---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Overview

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/dbtools)](https://CRAN.R-project.org/package=dbtools)
<!-- badges: end -->

`dbtools` provides helpers to insert, update, or delete in a database table with
the rows of a data frame. The main functions are:

* `db_insert_data()` to insert rows,
* `db_update_data()` to update rows
* `db_insert_missing_data()` to insert new rows,
* `db_upsert_data()` to insert new and update existing rows,
* `db_delete_data()` to delete rows.

While they are simple to use they also provide these features:

* _batch operations_ to increase the speed,
* _custom SQL code_ to have even better control,
* _nesting transactions_ with the help of `with_transaction()`,
* _returning_ the inserted/updated rows,
* _upserts/insert missing_ also possible without a unique index.

And of course they can easily be used with the pipe `%>%`.

## Installation

<!-- You can install the released version of dbtools from [CRAN](https://CRAN.R-project.org) with: -->

<!-- ``` r -->
<!-- install.packages("dbtools") -->
<!-- ``` -->

<!-- ### Development version -->
<!-- To get a bug fix, or use a feature from the development version, you can install dbtools from GitHub. -->

`dbtools` is not yet on CRAN and can only be installed from GitHub.

```{r eval=FALSE}
# install.packages("devtools")
devtools::install_github("mgirlich/dbtools")
```

## Insert (Missing) Data

Insert records

```{r}
library(dbtools)

if (DBI::dbExistsTable(con_memdb(), "dbtools")) {
  DBI::dbRemoveTable(con_memdb(), "dbtools")
}

dbplyr::db_copy_to(
  con = con_memdb(),
  table = "dbtools",
  values = data.frame(
    id = integer(),
    value = character(),
    update_counter = integer(),
    updated_at = character()
  )[0, ],
  overwrite = TRUE,
  unique_indexes = "id"
)

db_insert_data(
  data.frame(
    id = 1,
    value = c("a"),
    update_counter = 0,
    updated_at = Sys.time()
  ),
  table = "dbtools",
  con = con_memdb()
)

DBI::dbReadTable(con_memdb(), "dbtools")
```

If you want to insert data where the `id` is already in the table you get an
error because of the unique constraint on `id`:

```{r error=TRUE}
db_insert_data(
  data.frame(
    id = 1,
    value = c("a"),
    update_counter = 0,
    updated_at = Sys.time()
  ),
  table = "dbtools",
  con = con_memdb()
)
```

To only insert data where the `id` is not yet found in the table use
`db_insert_missing_data()`

```{r}
db_insert_missing_data(
  data = data.frame(
    id = 2:3,
    value = c("b", "c"),
    updated_at = Sys.time(),
    update_counter = c(2, 3)
  ),
  table = "dbtools",
  con = con_memdb(),
  conflict_target = "id"
)

DBI::dbReadTable(con_memdb(), "dbtools")
```

For more information on how to handle conflicts see the
[Conflicts and Unique Columns section](#conflicts).

## Update Data

```{r}
db_update_data(
  data = data.frame(
    id = 1:2,
    value = c("x", "y"),
    updated_at = Sys.time()
  ),
  table = "dbtools",
  con = con_memdb(),
  update = c("value", "updated_at"),
  where = "id"
)

DBI::dbReadTable(con_memdb(), "dbtools")
```

## Custom SQL

Let's say you only want to update rows with an `update_counter` of at most 2
and then you also want to increase the update counter by 1 for the updated rows.
This can easily be done by passing SQL code generated with `sql()`:

```{r}
db_update_data(
# sql_update(
  data = data.frame(
    id = 1:3,
    value = "z",
    updated_at = Sys.time()
  ),
  table = "dbtools",
  con = con_memdb(),
  update = list(
    "value",
    update_counter = sql("update_counter + 1"),
    "updated_at"
  ),
  where = list(
    "id",
    sql("update_counter <= 2")
  )
)

DBI::dbReadTable(con_memdb(), "dbtools")
```

## Data From Another Database Table

Instead of using local data to update/insert/upsert/delete you can also another
database table by providing its name to the `data` argument:

```{r}
if (DBI::dbExistsTable(con_memdb(), "dbtools2")) {
  DBI::dbRemoveTable(con_memdb(), "dbtools2")
}

dbplyr::db_copy_to(
  con = con_memdb(),
  table = "dbtools2",
  values = data.frame(
    dbtools_id = 1
  ),
  overwrite = TRUE
)

DBI::dbReadTable(con_memdb(), "dbtools")
DBI::dbReadTable(con_memdb(), "dbtools2")

db_delete_data(
  "dbtools2",
  "dbtools",
  con_memdb(),
  where = c(id = "dbtools_id")
)

DBI::dbReadTable(con_memdb(), "dbtools")
```

## Conflicts and Unique Columns {#conflicts}

When a table has a unique constraint on some columns (or it should have one but
for some reason it doesn't) and you try to insert data some rows might violate
this unique constraint. This is referred to as conflict target. There are two
types of conflict targets in `dbtools`:

* `sql_unique_cols()` to specify a set of unique columns,
* `sql_constraint()` to specify the name of an existing unique constraint.

Mind that `sql_constraint()` only works in newer versions of databases. If your
database version does not yet support the `ON CONFLICT` clause you have to use
`sql_unique_cols()` and use the `mode = "old"`.

## Generate SQL

You can also easily get the underlying SQL code with the corresponding
`sql_*()` function:

* `sql_insert()`
* `sql_insert_missing()`
* `sql_update()`
* `sql_delete()`

There are also some helper functions you might find useful:

* `sql_values()` to generate a `VALUES` clause from the data frame
