---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# dbtools

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/dbtools)](https://CRAN.R-project.org/package=dbtools)
<!-- badges: end -->

The goal of dbtools is to support

* inserting all or non existing data,
* updating existing data,
* upserting - i.e. inserting new and updating existing - data
* deleting data

in a database table using data from a dataframe.

## Installation

<!-- You can install the released version of dbtools from [CRAN](https://CRAN.R-project.org) with: -->

<!-- ``` r -->
<!-- install.packages("dbtools") -->
<!-- ``` -->

<!-- ### Development version -->
<!-- To get a bug fix, or use a feature from the development version, you can install dbtools from GitHub. -->

`dbtools` is not yet on CRAN and can only be installed from GitHub.

```{r eval=FALSE}
# install.packages("devtools")
devtools::install_github("mgirlich/dbtools")
```

## Example

### Database Operations

The database operations functions are all prefixed with `db_`. They follow the
idea that the first argument should be data, i.e. they can be piped. They all
support an optional returning clause, batchwise operations and control whether
the operation should be atomic (i.e. in a transaction).
The basic idea is to pass a character vector of column names but one can also
use plain SQL (generated with `SQL()`) for a deeper control.

* `db_update_data()` updates data as in the dataframe.
* `db_insert_missing_data()` inserts missing data from the dataframe.
* `db_upsert_data()` upserts (i.e. inserts new and updates existing data).
* `db_delete_data()`

* `batch_wise_db()`

### SQL Generators

Each database helper function has a function that generates the underlying SQL.
These functions (all named `sql_*()`) are also exported in case one needs to
modify the SQL. In addition there are some other helpers exported:

* `sql_insert()` is the workhorse for `sql_insert_missing()` and `sql_upsert()`.
* `sql_do_nothing()` and `sql_do_update()` to generate the conflict clause in
`sql_conflict_insert()`.
* `sql_values()` to generate a `VALUES` clause. 
* `sql_add_returning()` to add a returning clause if returning is not `NULL`.
